---
title: "Disease Interventions"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(malariaAtlas)
library(leaflet)
library(ggridges)
library(readxl)
library(plotly)
library(patchwork)

knitr::opts_chunk$set(
  fig.width=6,
  fig.asp=.9,
  out.width ="90%"
)

theme_set(theme_minimal() +theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete=scale_fill_viridis_d
```

```{r, echo = FALSE}
countries = c("Nigeria", "Democratic Republic of the Congo", "Uganda","Mozambique", "Côte d'Ivoire", "Niger", "Burkina Faso", "Mali", "Angola", "United Republic of Tanzania", "India", "Ghana", "Cameroon", "Rwanda", "Benin", "Malawi", "Kenya", "Guinea", "Algeria", "Argentina", "Iran", "Malaysia","Timor-Leste", "China", "El Salvador", "Cabo Verde")
```

```{r, echo = FALSE}
estimate_cases_deaths_df = 
  read_excel(
    "./data/population_at_risk_estimated_cases_deaths.xls", sheet = "apendix_f_data", range = "A2:J1037") %>% 
  janitor::clean_names() %>% 
  filter(country %in% countries) %>% 
  select(-cases_lower, -cases_upper, -deaths_lower, -deaths_upper) %>% 
  rename(
    cases = cases_point,
    deaths = deaths_point
  ) %>% 
  mutate(
    prevalence = cases / population_at_risk,
    mortality = deaths / population_at_risk
  )
```

```{r, echo=FALSE}
household_stat_df =
  read_excel(
    "./data/household_survey_STAT.xls", range = "A4:T44", col_names = c("country", "source", "% of households with at least one ITN", "% of households with at least one ITN for every two persons who stayed in the household the previous night", "% of households with IRS in last 12 months", "% of households with at least one ITN and/or IRS in the past 12 months", "% of householdswith at least one ITN for every two persons and/or IRS in the past 12 months", "% of population with access to an ITN", "% of population who slept under an ITN last night", "% of ITNs that were used last night", "% of pregnant women who slept under an ITN", "% of pregnant women who took 3+ doses of IPTp", "% of children <5 years who slept under an ITN", "% of children <5 years with moderate or severe anaemia", "% of children <5 yearswith a positive RDT", "% of children <5 years with a positive microscopy blood smear", "% of children <5 years with fever in last 2 weeksfor whom advice or treatment was sought", "% of children <5 years with fever in last 2 weekswho had blood taken from a finger or heel for testing", "% of children <5 years with fever in last 2 weeks who took antimalarial drugs", "% of children <5 years with fever in last 2 weeks who took an ACT among those who received any antimalarial")) %>% 
  janitor::clean_names() %>% 
  filter(
    country %in% countries
  ) %>% 
  pivot_longer(
    percent_of_households_with_at_least_one_itn:percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial,
    names_to = "preventative_measure",
    values_to = "percentage"
  ) %>% 
  drop_na(percentage) %>% 
  mutate(
    percentage = as.numeric(percentage),
    year = substr(source, 1, 4),
   year = as.integer(year),
   year_end = substr(source, 6, 7),
   year_end = case_when(
     str_detect(year_end, "[MDA][IH]") == TRUE ~ NA_character_,
     str_detect(year_end, "[MDA][IH]") == FALSE ~ year_end
  ),
  year_end = str_c("20", year_end),
  year_end = as.integer(year_end)
  ) %>% 
  select(-source)
```


## Prevention

Vector control is the main way to prevent and reduce malaria transmission. If coverage of vector control interventions within a specific area is high enough, then a measure of protection will be conferred across the community.

WHO recommends protection for all people at risk of malaria with effective malaria vector control. Two forms of vector control – insecticide-treated mosquito nets and indoor residual spraying – are effective in a wide range of circumstances.

### ITN

Sleeping under an insecticide-treated net (ITN) can reduce contact between mosquitoes and humans by providing both a physical barrier and an insecticidal effect. Population-wide protection can result from the killing of mosquitoes on a large scale where there is high access and usage of such nets within a community.


```{r, echo=FALSE}
itn_df = 
household_stat_df %>% 
  filter(preventative_measure %in% c("percent_of_households_with_at_least_one_itn", 
                                     "percent_of_pregnant_women_who_slept_under_an_itn",
                                    "percent_of_children_5_years_who_slept_under_an_itn",
                                    "percent_of_population_who_slept_under_an_itn_last_night")) %>% 
  pivot_wider(
    names_from = preventative_measure,
    values_from = percentage
  ) %>% 
  right_join(estimate_cases_deaths_df, by = c("country", "year")) %>% 
  drop_na(percent_of_households_with_at_least_one_itn) %>% 
  mutate(
    country = as.factor(country),
    country = fct_reorder(country, prevalence)
  ) 
```

```{r, echo=FALSE}
#Children under an ITN
children_itn = 
itn_df %>% 
  ggplot(aes(x = prevalence*100, y = percent_of_children_5_years_who_slept_under_an_itn)) +
  geom_point() + 
  labs(title = "ITN Usage: Children <5 Years Age",x = "Prevalence (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

#Pregnant women under an ITN
preg_women_itn = 
itn_df %>% 
  ggplot(aes(x = prevalence*100, y = percent_of_pregnant_women_who_slept_under_an_itn)) +
  geom_point() + 
  labs(title = "ITN Usage: Pregnant Women",x = "Prevalence (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

# Households under an ITN

household_itn = 
itn_df %>% 
  ggplot(aes(x = prevalence*100, y = percent_of_households_with_at_least_one_itn)) +
  geom_point() + 
  labs(title = "ITN Usage: Households",x = "Prevalence (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

# Population under an ITN

population_itn = 
itn_df %>% 
  ggplot(aes(x = prevalence*100, y = percent_of_population_who_slept_under_an_itn_last_night)) +
  geom_point() + 
  labs(title = "ITN Usage: Population",x = "Prevalence (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")
```

```{r, echo = FALSE, message = FALSE}
(population_itn + household_itn) / (preg_women_itn + children_itn)
```

```{r, echo = FALSE, message = FALSE}
fit_children_prev = lm(prevalence*100 ~ percent_of_children_5_years_who_slept_under_an_itn, data = itn_df) %>% 
  broom::tidy()

fit_preg_prev = lm(prevalence*100 ~ percent_of_pregnant_women_who_slept_under_an_itn, data = itn_df) %>% 
  broom::tidy()

fit_household_prev = lm(prevalence*100 ~ percent_of_households_with_at_least_one_itn, data = itn_df) %>% 
  broom::tidy()

fit_population_prev = lm(prevalence*100 ~ percent_of_population_who_slept_under_an_itn_last_night, data = itn_df) %>% 
  broom::tidy()

prev_fits = bind_rows(fit_children_prev, fit_preg_prev, fit_household_prev, fit_population_prev) %>% filter(term != "(Intercept)")

prev_fits %>% 
  knitr::kable()
```

```{r, echo=FALSE}
# Children under an ITN
children_itn_mort = 
itn_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_children_5_years_who_slept_under_an_itn)) +
  geom_point() + 
  labs(title = "ITN Usage: Children <5 Years Age",x = "Mortality (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

# Pregnant women under an ITN
preg_women_itn_mort = 
itn_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_pregnant_women_who_slept_under_an_itn)) +
  geom_point() + 
  labs(title = "ITN Usage: Pregnant Women",x = "Mortality (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

# Households under an ITN
household_itn_mort = 
itn_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_households_with_at_least_one_itn)) +
  geom_point() + 
  labs(title = "ITN Usage: Households",x = "Mortality (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

# Population under an ITN
population_itn_mort = 
itn_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_population_who_slept_under_an_itn_last_night)) +
  geom_point() + 
  labs(title = "ITN Usage: Population",x = "Mortality (%)", y= "% ITN Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")
```

```{r, echo = FALSE, message = FALSE}
(population_itn_mort + household_itn_mort) / (preg_women_itn_mort + children_itn_mort)
```

```{r, echo = FALSE, message = FALSE}
fit_children_mort = lm(mortality*100 ~ percent_of_children_5_years_who_slept_under_an_itn, data = itn_df) %>% 
  broom::tidy()

fit_preg_mort = lm(mortality*100 ~ percent_of_pregnant_women_who_slept_under_an_itn, data = itn_df) %>% 
  broom::tidy()

fit_household_mort = lm(mortality*100 ~ percent_of_households_with_at_least_one_itn, data = itn_df) %>%
  broom::tidy()

fit_population_mort = lm(mortality*100 ~ percent_of_population_who_slept_under_an_itn_last_night, data = itn_df) %>% 
  broom::tidy()

mort_fits = bind_rows(fit_children_mort, fit_preg_mort, fit_household_mort, fit_population_mort) %>% filter(term != "(Intercept)")

mort_fits %>% 
  knitr::kable()
```

### IRS

Indoor residual spraying (IRS) with insecticides is another powerful way to rapidly reduce malaria transmission. It involves spraying the inside of housing structures with an insecticide, typically once or twice per year. To confer significant community protection, IRS should be implemented at a high level of coverage.

Globally, IRS protection declined from a peak of 5% in 2010 to 2% in 2018, with decreases seen across all WHO regions, apart from the WHO Eastern Mediterranean Region. The declines in IRS coverage are occurring as countries switch from pyrethroid insecticides to more expensive alternatives to mitigate mosquito resistance to pyrethroids. 

```{r, echo = FALSE, message = FALSE}
irs_df = 
household_stat_df %>% 
  filter(preventative_measure %in% c("percent_of_households_with_irs_in_last_12_months",
                                "percent_of_households_with_at_least_one_itn_and_or_irs_in_the_past_12_months",
                                "percent_of_householdswith_at_least_one_itn_for_every_two_persons_and_or_irs_in_the_past_12_months")
  ) %>% 
  pivot_wider(
    names_from = preventative_measure,
    values_from = percentage
  ) %>% 
  right_join(estimate_cases_deaths_df, by = c("country", "year")) %>% 
  drop_na(percent_of_households_with_at_least_one_itn_and_or_irs_in_the_past_12_months) %>% 
  mutate(
    country = as.factor(country),
    country = fct_reorder(country, prevalence)
  ) 
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
a_irs_mort = 
irs_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_households_with_irs_in_last_12_months)) +
  geom_point() + 
  labs(title = "IRS Usage: Only IRS Last 12 Months",x = "Mortality (%)", y= "% IRS Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

a_irs_mort
```

```{r, echo = FALSE, message = FALSE}
fit_irs_mort = lm(mortality*100 ~ percent_of_households_with_irs_in_last_12_months, data = irs_df) %>% 
  broom::tidy() %>% filter(term != "(Intercept)")
fit_irs_mort %>% 
  knitr::kable()
```

### Antimalarial drugs

Antimalarial medicines can also be used to prevent malaria. For travellers, malaria can be prevented through chemoprophylaxis, which suppresses the blood stage of malaria infections, thereby preventing malaria disease. For pregnant women living in moderate-to-high transmission areas, WHO recommends intermittent preventive treatment with sulfadoxine-pyrimethamine, at each scheduled antenatal visit after the first trimester. Similarly, for infants living in high-transmission areas of Africa, 3 doses of intermittent preventive treatment with sulfadoxine-pyrimethamine are recommended, delivered alongside routine vaccinations.

Since 2012, WHO has recommended seasonal malaria chemoprevention as an additional malaria prevention strategy for areas of the Sahel sub-region of Africa. The strategy involves the administration of monthly courses of amodiaquine plus sulfadoxine-pyrimethamine to all children under 5 years of age during the high transmission season.

```{r, echo = FALSE, message = FALSE}
drug_df = 
household_stat_df %>% 
  filter(preventative_measure %in% c("percent_of_children_5_yearswith_a_positive_rdt",
                                     "percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_antimalarial_drugs",
                               "percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial",
                                     "percent_of_pregnant_women_who_took_3_doses_of_ip_tp")
  ) %>% 
  pivot_wider(
    names_from = preventative_measure,
    values_from = percentage
  ) %>% 
  right_join(estimate_cases_deaths_df, by = c("country", "year")) %>% 
  drop_na(percent_of_children_5_yearswith_a_positive_rdt) %>% 
  mutate(
    country = as.factor(country),
    country = fct_reorder(country, prevalence)
  ) 
```

```{r, echo = FALSE, message = FALSE}
c_drug_prev = 
drug_df %>% 
  ggplot(aes(x = prevalence*100, y = percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial)) +
  geom_point() + 
  labs(title = "% Children: Took ACT",x = "Prevalence (%)", y= "% Childre"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

c_drug_mort = 
drug_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial)) +
  geom_point() + 
  labs(title = "% Children: Took ACT",x = "Mortality (%)", y= "% Children"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

c_drug_prev + c_drug_mort

```





