---
title: "Disease Interventions"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(malariaAtlas)
library(leaflet)
library(ggridges)
library(readxl)
library(plotly)
library(patchwork)

knitr::opts_chunk$set(
  fig.width=6,
  fig.asp=.9,
  out.width ="90%"
)

theme_set(theme_minimal() +theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete=scale_fill_viridis_d
```

```{r, echo = FALSE}
countries = c("Nigeria", "Democratic Republic of the Congo", "Uganda","Mozambique", "Côte d'Ivoire", "Niger", "Burkina Faso", "Mali", "Angola", "United Republic of Tanzania", "India", "Ghana", "Cameroon", "Rwanda", "Benin", "Malawi", "Kenya", "Guinea", "Algeria", "Argentina", "Iran", "Malaysia","Timor-Leste", "China", "El Salvador", "Cabo Verde")
```

```{r, echo = FALSE}
estimate_cases_deaths_df = 
  read_excel(
    "./data/population_at_risk_estimated_cases_deaths.xls", sheet = "apendix_f_data", range = "A2:J1037") %>% 
  janitor::clean_names() %>% 
  filter(country %in% countries) %>% 
  select(-cases_lower, -cases_upper, -deaths_lower, -deaths_upper) %>% 
  rename(
    cases = cases_point,
    deaths = deaths_point
  ) %>% 
  mutate(
    prevalence = cases / population_at_risk,
    mortality = deaths / population_at_risk
  )
```

```{r, echo=FALSE}
household_stat_df =
  read_excel(
    "./data/household_survey_STAT.xls", range = "A4:T44", col_names = c("country", "source", "% of households with at least one ITN", "% of households with at least one ITN for every two persons who stayed in the household the previous night", "% of households with IRS in last 12 months", "% of households with at least one ITN and/or IRS in the past 12 months", "% of householdswith at least one ITN for every two persons and/or IRS in the past 12 months", "% of population with access to an ITN", "% of population who slept under an ITN last night", "% of ITNs that were used last night", "% of pregnant women who slept under an ITN", "% of pregnant women who took 3+ doses of IPTp", "% of children <5 years who slept under an ITN", "% of children <5 years with moderate or severe anaemia", "% of children <5 yearswith a positive RDT", "% of children <5 years with a positive microscopy blood smear", "% of children <5 years with fever in last 2 weeksfor whom advice or treatment was sought", "% of children <5 years with fever in last 2 weekswho had blood taken from a finger or heel for testing", "% of children <5 years with fever in last 2 weeks who took antimalarial drugs", "% of children <5 years with fever in last 2 weeks who took an ACT among those who received any antimalarial")) %>% 
  janitor::clean_names() %>% 
  filter(
    country %in% countries
  ) %>% 
  pivot_longer(
    percent_of_households_with_at_least_one_itn:percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial,
    names_to = "preventative_measure",
    values_to = "percentage"
  ) %>% 
  drop_na(percentage) %>% 
  mutate(
    percentage = as.numeric(percentage),
    year = substr(source, 1, 4),
   year = as.integer(year),
   year_end = substr(source, 6, 7),
   year_end = case_when(
     str_detect(year_end, "[MDA][IH]") == TRUE ~ NA_character_,
     str_detect(year_end, "[MDA][IH]") == FALSE ~ year_end
  ),
  year_end = str_c("20", year_end),
  year_end = as.integer(year_end)
  ) %>% 
  select(-source)
```

```{r, echo = FALSE, message = FALSE}
owi_mortality_rates_df_longer = 
  read_csv("./data/malaria-death-rates-by-age.csv") %>% 
  janitor::clean_names()  %>% 
  pivot_longer(
    deaths_malaria_sex_both_age_under_5_rate:deaths_malaria_sex_both_age_age_standardized_rate,
    names_to = "age_group",
    names_prefix = "deaths_malaria_sex_both_age_",
    values_to = "mortality_rate"
  )
```

## Mortality Rates by Age Group

In the fight against malaria, children under the age of 5 are disproportionately affected across the world. The chart below visualizes data  from the organization [our world in data](https://ourworldindata.org/grapher/malaria-death-rates-by-age) and shows the breakdown of death rates from malaria (measured as number of deaths per 100,000 individuals) by age category. 

```{r, echo = FALSE, message = FALSE, warning = FALSE}
owi_mortality_rates_df_longer =
owi_mortality_rates_df_longer %>% 
  group_by(year, age_group) %>%
  summarize(avg_mortality_rate = mean(mortality_rate)) %>% 
  mutate(
    age_group = factor(age_group, levels = c("under_5_rate","5_14_years_rate", "15_49_years_rate",
                                                  "50_69_years_rate","70_years_rate",
                                                  "age_standardized_rate","all_ages_rate"))
  ) 


ggplotly(ggplot(owi_mortality_rates_df_longer,aes(x=year, y = avg_mortality_rate, color = age_group)) + geom_point() + geom_line()  + theme(legend.position = "right" )+ labs(title = "Malaria Mortality Rates by Age Group", x = "Year", y = "Mortality Rate (%)", color = "Age Group"))
```

The following analysis will look into malaria prevention measures shown to be effective in managing malaria in the countries of interest that bear about 85% global burden of the disease. These countries are: Nigeria, Democratic Republic of the Congo, Uganda, Mozambique, Côte d'Ivoire, Niger, Burkina Faso, Mali, Angola, United Republic of Tanzania, India, Ghana, Cameroon, Rwanda, Benin, Malawi, Kenya, Guinea, Burundi.

## Prevention

Vector control is the main way to prevent and reduce malaria transmission. If coverage of vector control interventions within a specific area is high enough, then a measure of protection will be conferred across the community.

WHO recommends protection for all people at risk of malaria with effective malaria vector control. Two forms of vector control – insecticide-treated mosquito nets and indoor residual spraying – are effective in a wide range of circumstances.

### Insecticide-treated nets

Sleeping under an insecticide-treated net (ITN) can reduce contact between mosquitoes and humans by providing both a physical barrier and an insecticidal effect. 

The below analysis quantifies the impact of usage of bednets by children under 5 and how mortality rates for that age group can be improved.

__Key Variables__

* __Childhood mortality under age 5__: (`under_5_mortality_rate`) number of deaths per 100,000 individuals. The available data ranges from the years 1990 through 2017.

* __% of ITN Usage by children under 5__: (`use_of_insecticide_treated_bed_nets_percent_of_under_5_population`) % of children under 5 who sleep under a bednet. The available data ranges from the years 1999 - 2017.

__Data Sources__

Both variables came from separate datasets that were gathered via surveys and tracked globally across countries. The analysis used a combined dataset to understand the impact of bednets usage on childhood mortality. The team looked at multiple sources to access data available in a structured format and explored datasets from the [World Malaria Report 2019](https://www.who.int/publications/i/item/9789241565721), [UNICEF Data Warehouse](https://data.unicef.org/resources/data_explorer/unicef_f/?ag=UNICEF&df=GLOBAL_DATAFLOW&ver=1.0&dq=.MNCH_IPTP+MNCH_ITN+MNCH_ITN2+MNCH_ITNOWN+MNCH_ITNPREG+MNCH_MLRACT+MNCH_MLRCARE+MNCH_MLRDIAG..&startPeriod=2016&endPeriod=2020), and finally landed on using the [Our World in Data](https://ourworldindata.org/grapher/malaria-death-rates-by-age) since the dataset had the most bednets related data. Each of the datasets we explored did not have much data about bednets per country.

__Hypothesis Testing__

* __Ho__: Use of bednets by children under 5 has no impact to under 5 mortality rates
* __H1__: Use of bednets by children under 5 reduces mortality rates for children under 5

__Results__

Scatterplot of % ITN usage against that year's associated mortality rates, broken up by country. The dataset for bednets only included data for 12 out of 19 countries of interest. 

```{r, echo = FALSE, message = FALSE}
owi_mortality_rates_df = read_csv("./data/malaria-death-rates-by-age.csv") %>% 
  janitor::clean_names() %>% 
  rename(
    under_5_mortality_rate = deaths_malaria_sex_both_age_under_5_rate,
    all_ages_mortality_rate = deaths_malaria_sex_both_age_all_ages_rate,
    age_5_14_years_mortality_rate = deaths_malaria_sex_both_age_5_14_years_rate,
    age_15_49_years_mortality_rate = deaths_malaria_sex_both_age_15_49_years_rate,
    age_50_69_years_mortality_rate = deaths_malaria_sex_both_age_50_69_years_rate,
    age_70_years_mortality_rate = deaths_malaria_sex_both_age_70_years_rate,
    age_standardized_mortality_rate = deaths_malaria_sex_both_age_age_standardized_rate,
    country = entity
  ) %>% 
select(-code)

children_itn_df = read_csv("./data/children-sleeping-under-treated-bednet.csv") %>% 
  janitor::clean_names() %>% 
  rename(
     country = entity
  ) %>% 
  select(-code)

household_itn_df = read_csv("./data/households-with-at-least-one-itn.csv") %>% 
  janitor::clean_names() %>% 
  rename(
     country = entity
  ) %>% 
  select(-code)

complete_df =
  owi_mortality_rates_df %>% 
  left_join(children_itn_df, by = c("country","year")) %>% 
  left_join(household_itn_df, by = c("country","year")) %>% 
  filter(
    country %in% countries
  ) %>% 
  drop_na(use_of_insecticide_treated_bed_nets_percent_of_under_5_population)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
complete_df %>% 
ggplot(aes(x = under_5_mortality_rate, y = use_of_insecticide_treated_bed_nets_percent_of_under_5_population, color = country)) + geom_point() + stat_smooth(method = "lm") + facet_wrap(country ~.) + labs(title = "ITN Usage Among Children Under 5 and Malaria Mortality Rates", x = "Mortality Rates per 1000", y = "% ITN Usage", color = "Country")
```

```{r, echo = FALSE, message = FALSE}
fit_itn_owi = lm(under_5_mortality_rate ~ use_of_insecticide_treated_bed_nets_percent_of_under_5_population + country, data = complete_df) %>% 
  broom::tidy()

fit_itn_owi %>% 
  filter(term != "(Intercept)") %>% 
  knitr:::kable()
  
```

Based on the above analysis, you can see that ITN usage among children improves mortality rates by 4 children per 100,000 children when controlled for countries. We explored an interaction effect between country and use of ITNs, but found no statistically signficant effect, but the dataset might also not be well powered to detect that effect. India, Kenya and Rwanda were also countries that were shown to not have a statistically significant effect - most likely because of lack of data in those countries.

### Indoor residual spraying

Indoor residual spraying (IRS) with insecticides is another powerful way to rapidly reduce malaria transmission. It involves spraying the inside of housing structures with an insecticide, typically once or twice per year. To confer significant community protection, IRS should be implemented at a high level of coverage.

Globally, IRS protection declined from a peak of 5% in 2010 to 2% in 2018, with decreases seen across all WHO regions, apart from the WHO Eastern Mediterranean Region. The declines in IRS coverage are occurring as countries switch from pyrethroid insecticides to more expensive alternatives to mitigate mosquito resistance to pyrethroids. 

We accessed the WHO data collected about IRS spraying. Because this is a practice not used as much, there was not as much data available and our analysis was not broken down by country.

__Key Variables__

* __Mortality (%)__: `mortality` This variable was calculated from the WHO data which has population at risk per year per country and number of deaths attributed to malaria per year per country. 

* __% of Households with IRS in Last 12 Months__: `percent_of_households_with_irs_in_last_12_months` This variable was accessed from a collection of WHO datasets.  

The following graph shows a weak relationship between IRS spraying and mortality rates due to lack of data.
```{r, echo = FALSE, message = FALSE}
irs_df = 
household_stat_df %>% 
  filter(preventative_measure %in% c("percent_of_households_with_irs_in_last_12_months",
                                "percent_of_households_with_at_least_one_itn_and_or_irs_in_the_past_12_months",
                                "percent_of_householdswith_at_least_one_itn_for_every_two_persons_and_or_irs_in_the_past_12_months")
  ) %>% 
  pivot_wider(
    names_from = preventative_measure,
    values_from = percentage
  ) %>% 
  right_join(estimate_cases_deaths_df, by = c("country", "year")) %>% 
  drop_na(percent_of_households_with_at_least_one_itn_and_or_irs_in_the_past_12_months) %>% 
  mutate(
    country = as.factor(country),
    country = fct_reorder(country, prevalence)
  ) 
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
a_irs_mort = 
irs_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_households_with_irs_in_last_12_months)) +
  geom_point() + 
  labs(title = "IRS Usage: Only IRS Last 12 Months",x = "Mortality (%)", y= "% IRS Usage"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

a_irs_mort
```

```{r, echo = FALSE, message = FALSE}
fit_irs_mort = lm(mortality*100 ~ percent_of_households_with_irs_in_last_12_months, data = irs_df) %>% 
  broom::tidy() %>% filter(term != "(Intercept)")
fit_irs_mort %>% 
  knitr::kable()
```

### Antimalarial drugs

Antimalarial medicines can also be used to prevent malaria. For travellers, malaria can be prevented through chemoprophylaxis, which suppresses the blood stage of malaria infections, thereby preventing malaria disease. 

The below analysis shows the impact on mortality rates if children are treated with ACT as an antimalarial if they are infected.

```{r, echo = FALSE, message = FALSE}
drug_df = 
household_stat_df %>% 
  filter(preventative_measure %in% c("percent_of_children_5_yearswith_a_positive_rdt",
                                     "percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_antimalarial_drugs",
                               "percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial",
                                     "percent_of_pregnant_women_who_took_3_doses_of_ip_tp")
  ) %>% 
  pivot_wider(
    names_from = preventative_measure,
    values_from = percentage
  ) %>% 
  right_join(estimate_cases_deaths_df, by = c("country", "year")) %>% 
  drop_na(percent_of_children_5_yearswith_a_positive_rdt) %>% 
  mutate(
    country = as.factor(country),
    country = fct_reorder(country, prevalence)
  ) 
```

```{r, echo = FALSE, message = FALSE}
c_drug_prev = 
drug_df %>% 
  ggplot(aes(x = prevalence*100, y = percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial)) +
  geom_point() + 
  labs(title = "% Children: Took ACT",x = "Prevalence (%)", y= "% Children"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

c_drug_mort = 
drug_df %>% 
  ggplot(aes(x = mortality*100, y = percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial)) +
  geom_point() + 
  labs(title = "% Children: Took ACT",x = "Mortality (%)", y= "% Children"
  ) + theme(legend.position = "none") +
  stat_smooth(method = "lm")

c_drug_prev + c_drug_mort

```

As with the IRS data, there is not as much data collected around ACT usage among children. There is a weak relationship in improvements in prevelance and mortality with ACT usage.

```{r, echo = FALSE, message = FALSE}
fit_drug_mort = lm(prevalence*100 ~ percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial, data = drug_df) %>% 
  broom::tidy() %>% filter(term != "(Intercept)")
fit_irs_mort %>% 
  knitr::kable(caption = "Linear relationship: prevalence and ACT usage among children under 5")

fit_drug_mort = lm(mortality*100 ~ percent_of_children_5_years_with_fever_in_last_2_weeks_who_took_an_act_among_those_who_received_any_antimalarial, data = drug_df) %>% 
  broom::tidy() %>% filter(term != "(Intercept)")
fit_irs_mort %>% 
  knitr::kable(caption = "Linear relationship: mortality and ACT usage among children under 5")
```
