---
title: "Mosquito Species Exploration"
author: "Murrel Pereira"
date: "11/9/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(malariaAtlas)
library(leaflet)
library(ggridges)

knitr::opts_chunk$set(
  fig.width=6,
  fig.asp=.9,
  out.width ="60%"
)

theme_set(theme_minimal() +theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete=scale_fill_viridis_d
```

## Import

Bring in species dataset as a tibble. Instructions to bring in dataset is here: [Mapped Atlas] (https://cran.r-project.org/web/packages/malariaAtlas/vignettes/overview.html).

## One Country (Madagascar)

```{r, include=FALSE}
MMR_vec_data <- getVecOcc(country = "Madagascar") %>% 
  as_tibble()

```

### Understand species breakdown

```{r}

MMR_vec_data %>% 
  count(species) %>% 
  arrange(desc(n)) %>% 
  knitr::kable()

MMR_vec_data %>% 
  mutate(
    species = as.factor(species),
    species = fct_infreq(species)
  ) %>% 
  ggplot(aes(x=species, fill = species)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.text=element_text(size=5),) 
```

### How many papers is this data from?

```{r}

MMR_vec_data %>%
  count(citation) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  knitr::kable()
  
```


### Are the species spread out regionally?

```{r}

pal <- colorFactor("viridis", MMR_vec_data %>% pull(species_plain))

MMR_vec_data %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~longitude, ~latitude, radius = 1, color = ~pal(species_plain)) %>% 
  addLegend("bottomright", pal = pal, values = ~species_plain,
    title = "Mosquito Species",
    opacity = 0.8
  )

```

## Look across Africa

Import data
```{r, include=FALSE}
Africa_vec_data = 
  getVecOcc(continent = "Africa") %>% 
  as_tibble()
```

What is the spread of the different species across different countries - bar chart and scatterplot?

The bar chart is harder to read but tells you about the most predominant mosquito species and the scatterplot tells you a little more about prevalent species in different countries.
```{r}
Africa_vec_data %>% 
  mutate(
    species_plain = as.factor(species_plain),
    species_plain = fct_infreq(species_plain)
  ) %>% 
  ggplot(aes(x=species_plain, fill = country)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.text=element_text(size=5),
        legend.title=element_blank()) +
  guides(color = guide_legend(override.aes = list(size = 0.2))) +
  labs(title = "Count of Mosquito Species in African countries",x = "Mosquito Species"
  )

Africa_vec_data %>% 
  group_by(country, species_plain) %>% 
  summarise(n = n(), max_lat = max(latitude), max_long = max(longitude)) %>% 
  mutate(
    country = as.factor(country)
  ) %>% 
  ggplot(aes(x=fct_reorder(country,desc(n)), y=n, fill = species_plain, color = species_plain)) + geom_point() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
         
```

What is the density of different species across different countries?

```{r}
pal = colorFactor("viridis", Africa_vec_data %>% pull(species_plain))

Africa_vec_data %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~longitude, ~latitude, radius = 1, color = ~pal(species_plain)) %>% 
  addLegend("bottomleft", pal = pal, values = ~species_plain,
    title = "Mosquito Species",
    opacity = 0.5
  )

```

Summary data across Africa
```{r}
Africa_vec_data %>% 
  group_by(country, species_plain) %>% 
  summarise(n = n(), max_lat = max(latitude), max_long = max(longitude)) %>% 
  ggplot(aes(x=max_lat, y=max_long, color = species_plain, fill = species_plain, size = n)) + geom_point()
```

Turn above to a leaflet

```{r}

pal = colorFactor("viridis", Africa_vec_data %>% pull(species_plain))
Africa_vec_data %>% 
  group_by(country, species_plain) %>% 
  summarise(n = n(), max_lat = max(latitude), max_long = max(longitude)) %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~max_long, ~max_lat, radius = ~sqrt(n), color = ~pal(species_plain)) %>% 
  addLegend("bottomleft", pal = pal, values = ~species_plain,
    title = "Mosquito Species",
    opacity = 0.5
  )
```

# Do this across multiple continents (South America, Africa, and Asia)

Import data
```{r, include=FALSE}
Asia_vec_data = 
  getVecOcc(continent = "Asia") %>% 
  as_tibble()

SA_vec_data = 
  getVecOcc(continent = "Americas") %>% 
  as_tibble()

continent_vec_data = bind_rows(Africa_vec_data, Asia_vec_data, SA_vec_data)
```

Bar chart and scatterplot

```{r}
continent_vec_data %>% 
  mutate(
    species_plain = as.factor(species_plain),
    species_plain = fct_infreq(species_plain)
  ) %>% 
  ggplot(aes(x=species_plain, fill = continent_id)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.text=element_text(size=5),
        legend.title=element_blank()) +
  guides(color = guide_legend(override.aes = list(size = 0.2))) +
  labs(title = "Count of Mosquito Species in African, SA, Asian countries",x = "Mosquito Species"
  )

continent_vec_data %>% 
  group_by(country,continent_id, species_plain) %>% 
  summarise(n = n(), max_lat = max(latitude), max_long = max(longitude)) %>% 
  mutate(
    continent_id = as.factor(continent_id)
  ) %>% 
  ggplot(aes(x=max_long, y=n, fill = species_plain, color = species_plain)) + geom_point() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
         
```


What is the density of different species across different countries?

```{r}
pal = colorFactor("viridis", continent_vec_data %>% pull(species_plain))

continent_vec_data %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~longitude, ~latitude, radius = 1, color = ~pal(species_plain)) %>% 
  addLegend("bottomright", pal = pal, values = ~species_plain,
    title = "Mosquito Species",
    opacity = 0.5
  )

```

Summary data across continents
```{r}
continent_vec_data %>% 
  group_by(country, species_plain) %>% 
  summarise(n = n(), max_lat = max(latitude), max_long = max(longitude)) %>% 
  ggplot(aes(x=max_lat, y=max_long, color = species_plain, fill = species_plain, size = n)) + geom_point()
```

Turn above to a leaflet

```{r}

pal = colorFactor("viridis", continent_vec_data %>% pull(species_plain))
continent_vec_data %>% 
  group_by(country, species_plain) %>% 
  summarise(n = n(), max_lat = max(latitude), max_long = max(longitude)) %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~max_long, ~max_lat, radius = ~sqrt(n), color = ~pal(species_plain)) #%>% 
 # addLegend("bottomleft", pal = pal, values = ~species_plain,
  #  title = "Mosquito Species",
  #  opacity = 0.5
 # )
```

