---
title: "Disease Statistics"
output: 
  html_document:
    toc: true
    toc_float: true
---


While many countries in North America and parts of Europe have been able to eliminate the incidence of Malaria, many parts of the world are still plagued by this vector borne illness. The plots below show the changes in prevalence, incidence and mortality rates as observed over time, comparing the reported cases with the estimated cases for our countries of interest. Factors such as temperature, intervention , vector and parasite species etc contribute to the changes observed and influence these rates over time.



```{r message = FALSE, warning = FALSE, include=FALSE}
library(tidyverse)
library(rvest)
library(httr)
library(plotly)
library(ggplot2)
library(viridis)


burden = read_csv("./data/malariawho.csv") %>% 
  filter(country %in% c("Nigeria", "Democratic Republic of the Congo", "Uganda", "Mozambique", "Côte d'Ivoire", "Niger", "Burkina Faso", "Mali", "Angola", "United Republic of Tanzania", "India", "Ghana", "Cameroon", "Rwanda", "Benin", "Malawi", "Kenya", "Guinea", "Burundi"))
```


## Prevalence

Plots of prevalence reported and estimated by WHO by country over years. Interestingly, prevalence seems to increase in most countries over the years.

```{r message = FALSE, warning = FALSE, echo = FALSE}
prevalence = burden %>%
  select(country, year, reported_cases, estimated_cases, low_est_cases, high_est_cases) %>%
  plot_ly(x = ~year) %>% 
  add_trace(y = ~reported_cases, type = "scatter", mode = "lines", color = ~country, colors = "Reds", name = "Reported", hovetemplate = "%{country}") %>% 
   add_trace(y = ~estimated_cases, type = "scatter", mode = "marker", color = ~country, colors = "Reds", name = "Estimated", hoverinfo = "%{country}") %>% 
  layout(
    hovermode = "x unified", 
    showlegend = FALSE,
    title = "Reported and estimated number of cases",
    yaxis = list(title = "Cases"),
    updatemenus = list(
      list(
        type = "dropdown",
        y = 0.8,
        active = 0,
        buttons = list(
          list(method = "restyle",
               args = list("visible", c(T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Angola"),
          list(method = "restyle",
               args = list("visible", c(F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Benin"),
          list(method = "restyle",
               args = list("visible", c(F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burkina Faso"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burundi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Cameroon"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Côte d'Ivoire"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Democratic Republic of Congo"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Ghana"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F)), 
               label = "Guinea"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F)), 
               label = "India"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F)), 
               label = "Kenya"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F)), 
               label = "Malawi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F)), 
               label = "Mali"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F)), 
               label = "Mozambique"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F)), 
               label = "Niger"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F)), 
               label = "Nigeria"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F)), 
               label = "Rwanda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F)), 
               label = "Uganda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T)), 
               label = "United Republic of Tanzania")
          ))
    ))


prevalence
```


## Mortality

Mortality is documented by reported deaths and estimated deaths. As expected, reported deaths are far lower than estimated deaths.  However, from estimated deaths, we see some countries trending towards lower numbers!

```{r message = FALSE, warning = FALSE, echo = FALSE}
mortality = burden %>%
  select(country, year, reported_deaths, estimated_deaths, low_est_death, high_est_death) %>%
  plot_ly(x = ~year, color = ~country) %>% 
  add_trace(y = ~reported_deaths, type = "scatter", mode = "lines", color = ~country, colors = "Blues", name = "Reported") %>% 
  add_trace(y = ~estimated_deaths, type = "scatter", mode = "marker", color = ~country, colors = "Blues", name = "Estimated") %>% 
  layout(
    hovermode = "x unified", 
    showlegend = FALSE,
    title = "Reported and estimated deaths",
    yaxis = list(title = "Deaths"),
    updatemenus = list(
      list(
        type = "dropdown",
        y = 0.8,
        active = 0,
        buttons = list(
          list(method = "restyle",
               args = list("visible", c(T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Angola"),
          list(method = "restyle",
               args = list("visible", c(F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Benin"),
          list(method = "restyle",
               args = list("visible", c(F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burkina Faso"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burundi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Cameroon"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Côte d'Ivoire"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Democratic Republic of Congo"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Ghana"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F)), 
               label = "Guinea"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F)), 
               label = "India"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F)), 
               label = "Kenya"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F)), 
               label = "Malawi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F)), 
               label = "Mali"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F)), 
               label = "Mozambique"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F)), 
               label = "Niger"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F)), 
               label = "Nigeria"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F)), 
               label = "Rwanda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F)), 
               label = "Uganda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T)), 
               label = "United Republic of Tanzania")
          ))
    ))

mortality
```

## Incidence

Plot of incidence rates over years.  There seems to be a peak around 2010 in incidence rates.

```{r message = FALSE, warning = FALSE, echo = FALSE}
incidence = burden %>%
  select(country, year, incidence_per1000) %>%
  plot_ly(x = ~year, color = ~country) %>% 
  add_trace(y = ~incidence_per1000, type = "scatter", mode = "lines") %>% 
  layout(
    title = "Incidence Rates",
    yaxis = list(title = "Incidence per 1000 people"))

incidence
```

