---
title: "Disease Statistics"
output: 
  html_document:
    toc: true
    toc_float: true
---


While many countries in North America and parts of Europe have been able to eliminate the incidence of Malaria, many parts of the world are still plagued by this vector borne illness. The plots below show the changes in prevalence, incidence and mortality rates as observed over time, comparing the reported cases with the estimated cases for our countries of interest. Factors such as temperature, intervention , vector and parasite species etc contribute to the changes observed and influence these rates over time.



```{r message = FALSE, warning = FALSE, include=FALSE}
library(tidyverse)
library(rvest)
library(httr)
library(plotly)
library(ggplot2)
library(viridis)


burden = read_csv("./data/malariawho.csv") %>% 
  filter(country %in% c("Nigeria", "Democratic Republic of the Congo", "Uganda", "Mozambique", "C么te d'Ivoire", "Niger", "Burkina Faso", "Mali", "Angola", "United Republic of Tanzania", "India", "Ghana", "Cameroon", "Rwanda", "Benin", "Malawi", "Kenya", "Guinea", "Burundi"))
```


## Prevalence

Plots of reported and estimated cases by WHO by country over years. Interestingly, prevalence seems to increase in most countries over the years. Read more about the mosquitos and parasites that cause malaria in the [Understand Causes tab](pr_mosq.html)!

```{r message = FALSE, warning = FALSE, echo = FALSE}
prevalence = burden %>%
  select(country, year, reported_cases, estimated_cases, low_est_cases, high_est_cases) %>%
  plot_ly(x = ~year) %>% 
  add_trace(y = ~reported_cases, type = "scatter", mode = "lines", color = ~country, colors = "Reds", name = "Reported") %>% 
   add_trace(y = ~estimated_cases, type = "scatter", mode = "marker", color = ~country, colors = "Reds", name = "Estimated") %>% 
    style(visible = "legendonly") %>% 
  layout(
    hovermode = "x unified", 
    showlegend = FALSE,
    title = "Reported and estimated number of cases",
    yaxis = list(title = "Cases"),
    updatemenus = list(
      list(
        type = "dropdown",
        y = 0.8,
        active = 0,
        buttons = list(
          list(method = "restyle",
               args = list("visible", c(T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Angola"),
          list(method = "restyle",
               args = list("visible", c(F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Benin"),
          list(method = "restyle",
               args = list("visible", c(F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burkina Faso"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burundi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Cameroon"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "C么te d'Ivoire"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Democratic Republic of Congo"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Ghana"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F)), 
               label = "Guinea"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F)), 
               label = "India"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F)), 
               label = "Kenya"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F)), 
               label = "Malawi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F)), 
               label = "Mali"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F)), 
               label = "Mozambique"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F)), 
               label = "Niger"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F)), 
               label = "Nigeria"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F)), 
               label = "Rwanda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F)), 
               label = "Uganda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T)), 
               label = "United Republic of Tanzania")
          ))
    ))


prevalence
```


## Mortality

Mortality is documented by number of reported deaths and estimated deaths. As expected, reported deaths are far lower than estimated deaths.  However, from estimated deaths, we see some countries trending towards lower numbers! Look at trends between mortality and weather patterns in the [Correlations tab](weather_patterns.html)!

```{r message = FALSE, warning = FALSE, echo = FALSE}
mortality = burden %>%
  select(country, year, reported_deaths, estimated_deaths, low_est_death, high_est_death) %>%
  plot_ly(x = ~year, color = ~country) %>% 
  add_trace(y = ~reported_deaths, type = "scatter", mode = "lines", color = ~country, colors = "Blues", name = "Reported") %>% 
  add_trace(y = ~estimated_deaths, type = "scatter", mode = "marker", color = ~country, colors = "Blues", name = "Estimated") %>% 
    style(visible = "legendonly") %>% 
  layout(
    hovermode = "x unified", 
    showlegend = FALSE,
    title = "Reported and estimated deaths",
    yaxis = list(title = "Deaths"),
    updatemenus = list(
      list(
        type = "dropdown",
        y = 0.8,
        active = 0,
        buttons = list(
          list(method = "restyle",
               args = list("visible", c(T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Angola"),
          list(method = "restyle",
               args = list("visible", c(F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Benin"),
          list(method = "restyle",
               args = list("visible", c(F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burkina Faso"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burundi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Cameroon"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "C么te d'Ivoire"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Democratic Republic of Congo"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Ghana"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F)), 
               label = "Guinea"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F)), 
               label = "India"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F)), 
               label = "Kenya"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F)), 
               label = "Malawi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F)), 
               label = "Mali"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F)), 
               label = "Mozambique"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F)), 
               label = "Niger"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F)), 
               label = "Nigeria"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F)), 
               label = "Rwanda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F)), 
               label = "Uganda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T)), 
               label = "United Republic of Tanzania")
          ))
    ))

mortality
```

## Incidence

Plot of incidence rates over years.  There seems to be a peak around 2010 in incidence rates. Learn more about how interventions relate to incidence and death in the [Interventions tab](interventions.html)!

```{r message = FALSE, warning = FALSE, echo = FALSE}
incidence = burden %>%
  select(country, year, incidence_per1000) %>%
  plot_ly(x = ~year, color = ~country) %>% 
  add_trace(y = ~incidence_per1000, type = "scatter", mode = "lines") %>% 
  layout(
    title = "Incidence Rates",
    yaxis = list(title = "Incidence per 1000 people"))

incidence
```

## Death to Cases Ratio

Here we have death to case ratios by country. These plots show the number of deaths divided by the number of cases. 

```{r, message = FALSE, warning = FALSE, echo = FALSE}
ratio = burden %>%
  select(country, year, reported_deaths, estimated_deaths, reported_cases, estimated_cases) %>%
  filter(country %in% c("Nigeria", "Democratic Republic of the Congo", "Uganda", "Mozambique", "Cote d'Ivoire", "Niger", "Burkina Faso", "Mali", "Angola", "United Republic of Tanzania", "India", "Ghana", "Cameroon", "Rwanda", "Benin", "Malawi", "Kenya", "Guinea", "Burundi")) %>%
  mutate(
    est_ratio = (estimated_deaths / estimated_cases) * 100,
    rep_ratio = (reported_deaths / reported_cases) * 100
  ) %>% 
    plot_ly(x = ~year, color = ~country) %>% 
  add_trace(y = ~rep_ratio, type = "scatter", mode = "lines", color = ~country, colors = "Greens", name = "Reported %") %>% 
  add_trace(y = ~rep_ratio, type = "scatter", mode = "marker", color = ~country, colors = "Greens", name = "Estimated %") %>% 
    style(visible = "legendonly") %>% 
  layout(
    hovermode = "x unified", 
    showlegend = FALSE,
    title = "Death to case ratio",
    yaxis = list(title = "Deaths/Cases"),
    updatemenus = list(
      list(
        type = "dropdown",
        y = 0.8,
        active = 0,
        buttons = list(
          list(method = "restyle",
               args = list("visible", c(T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Angola"),
          list(method = "restyle",
               args = list("visible", c(F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Benin"),
          list(method = "restyle",
               args = list("visible", c(F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burkina Faso"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Burundi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Cameroon"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "C么te d'Ivoire"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Democratic Republic of Congo"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F,F)), 
               label = "Ghana"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F,F)), 
               label = "Guinea"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F,F)), 
               label = "India"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F,F)), 
               label = "Kenya"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F,F)), 
               label = "Malawi"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F,F)), 
               label = "Mali"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F,F)), 
               label = "Mozambique"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F,F)), 
               label = "Niger"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F,F)), 
               label = "Nigeria"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F,F)), 
               label = "Rwanda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T,F)), 
               label = "Uganda"),
          list(method = "restyle",
               args = list("visible", c(F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,F,T)), 
               label = "United Republic of Tanzania")
          ))
    ))

ratio
```

