---
title: "Weather Patterns"
author: "Team Malaria"
date: "11/10/2020"
output: github_document
---

```{r setup}
library(tidyverse)
library(rvest)
library(httr)
library(plotly)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6,
  out.width = "90%")
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_color_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Import Data from Kaggle 
```{r}
library(readr)
library(lubridate)
library(tidyr)
library(dplyr)
library(ggplot2)
library(viridis)
library(tidyverse)
require(gridExtra)

```

#Import and Tidy data for Global Temperature per country:
```{r}
Globalcountry_temp_df =  read_csv("./Data/GlobalLandTemperaturesByCountry.csv") %>%
  separate(dt, c("Year","Month", "Day")) %>%
  drop_na() 

# Outbreak from 1750-1755
outbreak_df = Globalcountry_temp_df %>%
  filter(Year > 1750, Year < 1755)


# No outbreak from 2000-2005
nonoutbreak_df = Globalcountry_temp_df %>%
  filter(Year > 2000, Year < 2005)
  
  #mutate year, month, day into numeric even though categorical data?
```


## Visualiations 

```{r Malaria_outbreak_df}

outbreak_df_plot = outbreak_df %>%
  ggplot(aes(x = Month, y = AverageTemperature, color = Country)) + 
  geom_point(alpha = 0.25) + geom_line() +
  coord_cartesian() +
  theme(legend.position = "bottom") +
  labs(
    title = "Average Temperature per country over each month for 1750-1755",
    x = "Month",
    y = "Average Temperature ")
outbreak_ggplotly = ggplotly(outbreak_df_plot)



nonoutbreak_plot = nonoutbreak_df %>%
  ggplot(aes(x = Month, y = AverageTemperature, color = Country)) + 
  geom_point(alpha = 0.25) + geom_line() +
  coord_cartesian() +
  theme(legend.position = "bottom") +
  labs(
    title = "Average Temperature per country over each month for 2000-2005",
    x = "Month",
    y = "Average Temperature ")
nonoutbreak_ggplotly = ggplotly(nonoutbreak_plot)

grid.arrange(outbreak_df_plot, nonoutbreak_plot, ncol = 2)


##Calculate the Uncertainty of the data sets

#Outbreak_uncertainty_plot = outbreak_df %>%
     #group_by(Year) %>%
     #summarise(Year = first('Year'), 
     #avg_temp = mean(AverageTemperature, na.rm = T), 
     #avg_conf = mean(AverageTemperatureUncertainty, na.rm = T)) %>%
     #ggplot(mapping = aes(x = Year, y = avg_temp)) +
     #geom_line(color = viridis(1, begin = 0.7), mapping = aes(y = avg_conf)) +
     #theme_bw()

#Nonoutbreak_uncertainty_plot = nonoutbreak_df %>%
     #group_by(Year) %>%
     #summarise(Year = first(Year), 
     #avg_temp = mean(AverageTemperature, na.rm = T), 
     #avg_conf = mean(AverageTemperatureUncertainty, na.rm = T)) %>%
     #ggplot(mapping = aes(x = Year, y = avg_temp)) +
     #geom_line(color = viridis(1, begin = 0.7), mapping = aes(y = avg_conf)) +
     #theme_bw()

```


# By season for outbreak
```{r outbreak_df}
make_seasons = function(m) {
  factor((m %% 12) %/% 3, labels = c('Winter', 'Spring', 'Summer', 'Autumn'))
 }
 
 seasons1_df = outbreak_df %>%
   mutate(Year = as.numeric(Year)) %>%
   mutate(Month = as.numeric(Month)) %>%
   mutate(Seasons = make_seasons(Month)) %>%
     group_by(Month, Seasons,Year) %>%
    summarise(Month = first(Month),    #what happens if remove first(Month)
               avg_temp = mean(AverageTemperature, na.rm = T), 
               avg_conf = mean(AverageTemperatureUncertainty, na.rm = T)) 


 outbreakseasons_plot = seasons1_df %>%
  ggplot(mapping = aes(x = Year, y = avg_temp, color = Seasons)) + 
  geom_point() +
  geom_line(mapping = aes(color = Seasons)) + 
  theme(legend.position = "bottom") +
  scale_color_viridis(discrete = T) +
  labs(
    title = "Average Temperature per Seaons for 1750-1755",
    x = "Year",
    y = "Average Temperature ")

```

# By season for nonoutbreak
```{r nonoutbreak_df}

make_seasons = function(m) {
  factor((m %% 12) %/% 3, labels = c('Winter', 'Spring', 'Summer', 'Autumn'))
 }
 
 seasons2_df = nonoutbreak_df %>%
   mutate(Year = as.numeric(Year)) %>%
   mutate(Month = as.numeric(Month)) %>%
   mutate(Seasons = make_seasons(Month)) %>%
     group_by(Month, Seasons, Year) %>%
    summarise(Month = first(Month),    #what happens if remove first(Month)
               avg_temp = mean(AverageTemperature, na.rm = T), 
               avg_conf = mean(AverageTemperatureUncertainty, na.rm = T)) 



 nonoutbreakseasons_plot = seasons2_df %>%
  ggplot(mapping = aes(x = Year, y = avg_temp, color = Seasons)) + 
  geom_point() +
  geom_line(mapping = aes(color = Seasons)) + 
  theme(legend.position = "bottom") +
  scale_color_viridis(discrete = T) +
  labs(
    title = "Average Temperature per Seaons for 2000-2005",
    x = "Year",
    y = "Average Temperature ")
 
```


## Leaflet Visualizations???



Describe the data set:

access from: (http://berkeleyearth.org/archive/data/):
Repackaged data from a newer compilation put together by the Berkeley Earth, which is affiliated with Lawrence Berkeley National Laboratory. The Berkeley Earth Surface Temperature Study combines 1.6 billion temperature reports from 16 pre-existing archives. The dataset includes several files:

The Globalcountry_temp_df comes from the Global Land Temperatures by Country (GlobalLandTemperaturesByCountry.csv) and contains the following:

Year: starts in 1850 for max and min land temperatures and land temperatures
Month, Day, Average Temperature, temperature, country 
Average Temperature Uncertainty: the 95% confidence interval around the average 



